---
title: Graphical Queries - Click Events
layout: post
output: 
  md_document:
    preserve_yaml: true
---

_An introduction to click events in Shiny_

[Code](https://github.com/krisrs1128/stat679_code/blob/main/notes/week3-1.Rmd), [Recording](https://mediaspace.wisc.edu/media/Week%207%20-%201%3A%20Time%20Series%20Visualization%20(I)/1_dubryjey)

```{r, echo = FALSE}
library(knitr)
opts_knit$set(base_dir = "/", base.url = "/")
opts_chunk$set(
  warning = FALSE, 
  message = FALSE,
  fig.path = "stat679_notes/assets/week3-1/"
  )
```

1. Some of the most sophisticated interactive data visualizations are based on
the idea that user queries can themselves be defined visually. For example, to
select a date range, we could directly interact with a time series plot, rather
than relying on a slider input. Or, instead of a long dropdown menu of items, a
user could select items by clicking on bars in a bar plot. There are many
variations of this idea, but they all leverage graphical (rather than textual)
displays to define queries. The advantage of this approach is that it increases
information density -- the selection inputs themselves encode data.

1. To implement this in Shiny, we first need a way of registering user
interactions on plots themselves. We will consider two types of plot interaction
mechanisms: clicks and brushes. These can be specified by adding `click` or
`brush` events to `plotOutput` objects. 

1. This creates a UI with a single plot on which we will be able to track user
clicks,
    
    ```{r, eval = FALSE}
    ui <- fluidPage(
      plotOutput("plot", click = "plot_click")
    )
    ```
    
    Here, `plot_click` is an ID that can be used as `input$plot_click` in the
    server. We could name it however we want, but we need to be consistent
    across the UI and server (just like ordinary, non-graphical inputs).
  
1. Before, we just needed to place the `input$id` items within `render` and
`reactive` server components, and the associated outputs would automatically
know to redraw each time the value of any input was changed. Clicks are treated
slightly differently. We have to both (a) recognize when a click event has
occurred and (b) extract relevant information about what the click was referring
to. 

1. For (a), we generally use `observeEvent`,

    ```{r, eval = FALSE}
    observeEvent(
      input$plot_click,
      ... things to do when the plot is clicked ...
    )
    ```

    This piece of code will be run anytime the plot is clicked.
  
1. For (b), we can use the `nearPoints` helper function. Suppose the plot was
made using the data.frame `x`. Then

    ```{r, eval = FALSE}
    nearPoints(x, input$click)
    ```

    will return the samples in `x` that are close to the clicked location. We
    will often use a variant of this code that doesn't just return the closeby
    samples -- it returns all samples, along with their distance from the
    clicked location,
    
    ```{r, eval = FALSE}
    nearPoints(x, input$click, allRows = TRUE, addDist = TRUE)
    ```
    
1. We are almost ready to build a visualization whose outputs respond to
graphical queries. Suppose we want a scatterplot where point sizes update
according to their distance from the user's click. Everytime the plot is
clicked, we need to update the set of distances between samples and the clicked
point. We then need to rerender the plot to reflect the new distances. This
logic is captured by the block below,

    ```{r, eval = FALSE}
    server <- function(input, output) {
      dist <- reactiveVal(rep(1, nrow(x)))
      observeEvent(
        input$plot_click,
        dist(reset_dist(x, input$plot_click))
      )
      
      output$plot <- renderPlot({
        scatter(x, dist())
      })
    }
    ```
    
    The code above uses one new concept, the `reactiveVal` on the first line of
    the function. It is a variable that doesn't directly depend on any inputs,
    which can become a source node for downstream `reactive` and `render` nodes
    in the reactive graph. Anytime the variable's value is changed, all
    downstream nodes will be recomputed. A very common pattern is use an
    `observeEvent` to update a `reactiveVal` every time a graphical query is
    performed. Any plots that depend on this value will then be updated. For
    example,
    
    ```{r, eval = FALSE}
    val <- reactiveVal(initial_val) # initialize the reactive value
    
    observeEvent(
      ...some input event...
      ...do some computation...
      val(new_value) # update val to new_val
    )
    
    # runs each time the reactiveVal changes
    renderPlot({
      val() # get the current value of the reactive value
    })
    ```
        
1. So, revisiting the `dist` in the earlier code block, we see that it is
initialized as a vector of `1`'s whose length is equal to the number of rows of
`x`. Everytime the plot is clicked, we update the value of `dist` according to
the function `reset_dist`. Finally, the changed value of `dist` triggers a rerun
of `renderPlot`. Let's look at the full application in action. It makes a
scatterplot using the cars dataset and resizes points every time the plot is
clicked.

    ```{r code=readLines("https://raw.githubusercontent.com/krisrs1128/stat479_s22/main/_posts/2022-02-16-week05-01/apps/app1.R"), eval = FALSE}
    ```
    
    <iframe src="https://data-viz.it.wisc.edu/content/17a54e4c-5237-4018-b6fe-20c1555cf337/" allowfullscreen="" data-external="1" height=400 width=600></iframe>
    
1. The `reset_dist` function uses `nearPoints` to compute the distance between
each sample and the plot, each time the plot is clicked. The associated reactive
value `dist` gets changed, which triggers `scatterplot` to run, and it is
encoded using size in the downstream ggplot2 figure.

1. We can make the plot more interesting by outputting a table showing the
original dataset. Using the same `dist()` call, we can sort the table by
distance each time the plot is clicked.

    ```{r code=readLines("https://raw.githubusercontent.com/krisrs1128/stat479_s22/main/_posts/2022-02-16-week05-01/apps/app2.R"), eval = FALSE}
    ```
    
    <iframe src="https://data-viz.it.wisc.edu/content/9fcf3583-a74f-4d84-8d7c-68ac3595bf6b/" allowfullscreen="" data-external="1" height=750 width=800></iframe>

