---
title: Frequency Framing
layout: post
output: 
  md_document:
    preserve_yaml: true
---

_More evocative uncertainty visualization_

[Code](https://github.com/krisrs1128/stat679_code/blob/main/notes/week13-1.Rmd), [Recording](https://mediaspace.wisc.edu/media/Week%2013%20-%201%3A%20Frequency%20Framing/1_ywcaurej)

```{r, echo = FALSE}
library(knitr)
opts_knit$set(base_dir = "/", base.url = "/")
opts_chunk$set(
  warning = FALSE, 
  message = FALSE,
  fig.path = "stat679_notes/assets/week13-1/"
  )
```

```{r}
library(tidyverse)
library(ggdist)
library(distributional)
library(patchwork)
library(ungeviz)
theme_set(theme_bw())
```

1. Whenever we analyze a dataset, it's worth keeping in mind that our
observations could have turned out differently. Different participants might
have enrolled in the study, the sensor might be affected by measurement noise,
the author might have chosen a different word. Consequently, all inferences we
draw from data are accompanied by some degree of uncertainty. One of the basic
ideas of statistics is that we can precisely measure and honestly communicate
this uncertainty.

1. Yet, for most of this class, we've simply displayed the data as they are.
When most of us see a visualization like this, we don’t automatically wonder
about the hypothetical alternatives. Therefore, if we want our visual designs to
communicate uncertainty, we need clear ways of representing them graphically.
This will be our subject for the last few weeks.

1. For events with discrete outcomes (e.g., win the lottery or not), it is
possible to visualize the probabilities directly. For example, if there are just
two outcomes, we can plot the probability on along a line; if there are multiple
outcomes, a stacked bar chart can achieve the same effect. 

    ```{r, fig.width = 6, fig.height = 3}
    sample_grid <- function(p = 0.5) {
      expand.grid(seq_len(25), seq_len(25)) %>%
        mutate(response = sample(0:1, n(), replace = TRUE, prob = c(1 - p, p)))
    }
    
    true_prob <- c(0.01, 0.1, 0.4)
    outcomes <- map(true_prob, ~ sample_grid(.)) %>%
      bind_rows(.id = "p") %>%
      mutate(true_prob = true_prob[as.integer(p)])
    
    outcomes %>%
      group_by(p) %>%
      summarise(empirical_prob = mean(response)) %>%
      ggplot() +
        geom_col(aes(as.factor(true_prob), empirical_prob)) +
        scale_y_continuous(expand = c(0, 0, 0, 0.1)) +
        labs(x = "True Probability", y = "Empirical Probability")
    ```

1. However, there is experimental evidence that humans are poor at judging
probabilities represented in this way. A more effective alternative is frequency
framing. The idea is simple: create marks that represent the potential outcomes,
and ensure the frequency of the types reflects the underlying probabilities. For
example, the displays below represent the probabilities of events with 1, 10,
and 40% probability of occurring.

    ```{r, fig.width = 6, fig.height = 2}
    ggplot(outcomes) +
      geom_tile(aes(Var1, Var2, fill = as.factor(response)), col = "white", size = 0.6) +
      facet_wrap(~ true_prob) +
      scale_y_continuous(expand = c(0, 0)) +
      scale_x_continuous(expand = c(0, 0)) +
      coord_fixed() +
      labs(fill = "Outcome", col = "Outcome") +
      scale_fill_manual(values = c("#D9D9D9", "#595959")) +
      theme(
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank()
      )
    ```
  
1. How can we generalize this idea to continuous outcomes? For example, in an
upcoming election, party A might defeat party B by X%. We might believe that X
is distributed normally with mean 1.1 and standard deviation 2.4. One approach
is to visualize this using the density plot below.

    ```{r, fig.width = 3, fig.height = 1.5, dpi=300, echo = FALSE}
    estimate <- tibble(mean = c(1.1), sd = c(2.4))
    ggplot(estimate) +
      stat_slab(aes(xdist = dist_normal(mean, sd), y = 0, fill = after_stat(x > 0))) +
      scale_y_continuous(expand = c(0, 0, 0, 0.1)) +
      scale_fill_manual(values = c("#400D3C", "#ADD9CC")) +
      labs(x = "Forecasted Percentage Difference") +
      theme(
        axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position = "none",
        panel.grid = element_blank(),
        panel.border = element_blank()
      )
    ```
  
1. We created this density using the `ggdist` package with the `dist_normal`
distribution aesthetic. We split the colors between parties using the `fill`
argument.

    ```{r, eval = FALSE}
    estimate <- tibble(mean = c(1.1), sd = c(2.4))
    ggplot(estimate) +
      stat_slab(aes(xdist = dist_normal(mean, sd), y = 0, fill = after_stat(x > 0))) +
      scale_y_continuous(expand = c(0, 0, 0, 0.1)) +
      scale_fill_manual(values = c("#400D3C", "#ADD9CC")) +
      labs(x = "Forecasted Percentage Difference") +
      theme(
        axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position = "none",
        panel.grid = element_blank(),
        panel.border = element_blank()
      )
    ```
    
1. This visualization suffers from the same limitations as directly representing
probabilities in the discrete outcome case. The frequency framing version
creates a dotplot for hypothetical outcomes at locations proportional to their
probability. It seems like a simple change, but it is often better perceived,
since judging counts is often easier than comparing areas.

    ```{r, fig.width = 3, fig.height = 1.5, dpi=300, echo = FALSE}
    ggplot(estimate) +
      stat_dots(aes(xdist = dist_normal(mean, sd), y = 0, fill = after_stat(x > 0)), col = "white") +
      scale_y_continuous(expand = c(0, 0, 0, 0.1)) +
      scale_fill_manual(values = c("#400D3C", "#ADD9CC")) +
      labs(x = "Forecasted Percentage Difference") +
      theme(
        axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position = "none",
        panel.grid = element_blank(),
        panel.border = element_blank()
      )
    ```
  
1. We created this dotplot using the `geom_dotplot` layer from `ggdist`. Like in
ordinary ggplot2, ggdist allows us to compose layers. For example, we can create
a combined dot and density plot (this is called a rain cloud plot).
    
    ```{r, eval = FALSE}
    ggplot(estimate) +
      stat_dots(aes(xdist = dist_normal(mean, sd), y = 0, fill = after_stat(x > 0)), col = "white") +
      scale_y_continuous(expand = c(0, 0, 0, 0.1)) +
      scale_fill_manual(values = c("#400D3C", "#ADD9CC")) +
      labs(x = "Forecasted Percentage Difference") +
      theme(
        axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position = "none",
        panel.grid = element_blank(),
        panel.border = element_blank()
      )
    ```
  
1. We can compactly display the uncertainties for many inferences by stacking
dotplots or intervals. For example, in the plot below, we show chocolate quality
across many countries. More than simply describing the average quality of
chocolate bars across countries, it helps answer the more nuanced question, "If
I were in a grocery store and randomly picked a chocolate bar from country X,
what would I expect its quality to be?" That said, it can still be helpful to
include summary intervals showing the mean +/- 1 and two standard deviations
(right plot).

    ```{r, fig.height = 8, fig.width = 13}
    cacao_small <- cacao %>%
      filter(location %in% c("Ecuador", "Australia", "Belgium", "Switzerland", "Germany"))
    
    p <- ggplot(cacao_small, aes(rating, reorder(location, rating))) + theme(axis.title.y = element_blank())
    p1 <- p + stat_dots(dotsize = .3, binwidth = .2, fill = "#6894A6")
    p2 <- p + stat_dotsinterval(dotsize = .3, binwidth = .2, fill = "#6894A6")
    
    p1 + p2
    ```
    
1. A variation of stacked dotplots are called ridgeline plots. This no longer
uses dots to support frequency framing, but it's much easier to view partially
overlapping density plots than dot plots. 

    ```{r}
    p + stat_slab(height = 2, col = "black",  fill = "#6894A6")
    ```
    
    Since these are densities, it also isn't important for the different countries
to have approximately the same counts. When working with a larger subset of
countries, this makes a big difference.

    ```{r, fig.height = 6, fig.width = 10}
    data(cacao)
    p <- cacao %>%
      group_by(location) %>%
      filter(n() > 12) %>%
      ggplot(aes(rating, reorder(location, rating))) +
      theme(axis.title.y = element_blank())
        
    p1 <- p + 
      stat_slab(height = 2, col = "black", fill = "#6894A6") +
      labs(title = "Legible :)")
    p2 <- p + 
      stat_dots(dotsize = 0.3, binwidth = 0.2, col = "#6894A6") +
      labs(title = "Illegible :(")
    p1 + p2
    ```

1. Frequency framing is a reasonable default strategy for representing
uncertainty. However, there are many other ways to visually encode distributions
over outcomes, and they can be worth knowing for more specialized visualization
problems. For example, if we had many more countries in our chocolate bar
ratings plot, we need a more compact way of representing uncertainty. Our next
few notes explore alternatives for this and other situations.
