---
title: Geospatial Visualization (I)
layout: post
output: 
  md_document:
    preserve_yaml: true
---

_An introduction to visualizing geospatial data_

[Code](https://github.com/krisrs1128/stat679_code/blob/main/notes/week7-3.Rmd), [Recording](https://mediaspace.wisc.edu/media/Week%207%20-%203%3A%20Geospatial%20Visualization%20(I)/1_nimg6uzj)

```{r, echo = FALSE}
library(knitr)
opts_knit$set(base_dir = "/", base.url = "/")
opts_chunk$set(
  warning = FALSE, 
  message = FALSE,
  fig.path = "stat679_notes/assets/week7-3/"
)
```

```{r}
library(sf)
library(tmap)
library(terra)
library(ceramic)
library(tidyverse)
```

1. In the same way that time plays a central role in analysis of time series
datasets, space / location is often a central element in geographic data
visualization. Indeed, these two types of visualization have much longer
histories than even the simple scatterplot (which only emerged in the 19th
century, as the abstract analog of map).

1. There are two types of geographic data worth distinguishing between — vector
and raster data. Vector data formats are used to store geometric information,
like the locations of hospitals (points), trajectories of bus routes (lines), or
boundaries of counties (polygons). It’s useful to think of the associated data
as being spatially enriched data frames, with each row corresponding to one of
these geometric features. Vector data are usually stored in .geojson, .wkt,
.shp, or .topojson formats.

1. Raster data give a measurement along a spatial grid. You can think of them as
spatially enriched matrices, where the metadata says where on the earth each
entry of the matrix is associated with. Raster data are often stored in .tiff
format.

1. There are three packages that are especially useful for working with spatial
data in R. `sf` and `terra` are designed to support manipulation of vector and
raster data, respectively. `tmap` is a visualization package applicable to
either format (or even combinations of them).

1. Let’s look at some vector data. The examples below read in a road network
from Brasilia (multilines), glacier boundaries in Nepal (multipolygons), and
locations of hospitals in Madison (multipoints). Notice that each row is
georeferenced. Try searching the lat / lon coordinates online, [like this](https://goo.gl/maps/dddPgZwDWnYUg5t86).

    ```{r}
    roads <- read_sf("https://raw.githubusercontent.com/krisrs1128/stat679_code/main/examples/week7/week7-3/brasilia_small.geojson")
    roads
    
    clinics <- read_sf("https://raw.githubusercontent.com/krisrs1128/stat679_code/main/examples/week7/week7-3/clinics.geojson")
    clinics
    
    glaciers <- read_sf("https://raw.githubusercontent.com/krisrs1128/stat679_code/main/examples/week7/week7-3/glaciers.geojson")
    glaciers
    ```
    
  

1. We will use `tmap` to visualize each of these datasets. The logic of the
package mirrors that of `ggplot2`, we use `tmap_shape` to tell the package our
source data, and then we can add on layers that add visual marks based on it.
For example, to draw the road network, we use add on a `tmap_line` layer. For
the glaciers and hospitals, we can use the analogous fill and point layers.

    ```{r}
    tm_shape(roads) +
      tm_lines()
    
    tm_shape(clinics) +
      tm_dots(size = 1, col = "red")
    
    tm_shape(glaciers) +
      tm_polygons()
    ```
  
1. Remember that each element of a vector dataset can be thought of as a row of
a data.frame (just with added georeferencing information). This means that it is
often possible to encode features of each element using properties of the marks
used to represent them. For example, we’ve shaded in each of the glaciers by
their thickness, which shows that the larger and more central ones tend to be
thicker. These kinds of fill-encoded polygon visualizations are sometimes called
choropleths.

    ```{r}
    tm_shape(glaciers) +
      tm_borders() +
      tm_fill(col = "Thickness", palette = "Blues", legend.hist = TRUE) +
      tm_layout(legend.outside = TRUE)
    ```
  
1. What about raster data? We can again use `tmap`, but we have to use
different types of layers. For example, the data read in below come from the
Sentinel 2 satellite, an open (and continually updated) collection of satellite
images maintained by the European Space Agency. They give imagery associated
with one small subset of the glaciers labels. To visualize just one image
channel from this data, we can use `tmap_raster`.
    
    ```{r}
    im <- rast("https://uwmadison.box.com/shared/static/lpmujy5odtt3otpq1bluiv16fkg0o4t2.tif") %>%
      stretch()
    
    # only the first layer
    tm_shape(subset(im, 1)) +
      tm_raster()
    ```
  
1. Many raster datasets have more than just one channel (our Sentinel dataset
has 13). We can plot combinations of them using `tmap_rgb`. For example,
channels 2 - 4 correspond to the usual RGB channels that appear in ordinary
camera images.

    ```{r}
    tm_shape(im) +
      tm_rgb()
    ```

1. We can overlay vector data over a raster basemap by combining layers. We are
retrieving the basemap using the `cc_location` function in `ceramic`.

    ```{r}
    # you can get your own at https://account.mapbox.com/access-tokens/
    Sys.setenv(MAPBOX_API_KEY="pk.eyJ1Ijoia3Jpc3JzMTEyOCIsImEiOiJjbDYzdjJzczQya3JzM2Jtb2E0NWU1a3B3In0.Mk4-pmKi_klg3EKfTw-JbQ")
    basemap <- cc_location(loc= c(-89.401230, 43.073051), buffer = 15e3)
    
    tm_shape(basemap) +
      tm_rgb() +
      tm_shape(clinics) +
      tm_dots(col = "red", size = 1)
    ```
  
1. In all the discussion above, we’ve worked with the original spatial data
that were given to us. In practice, we will often want to manipulate these data
before making a final visualization — for example, we may want to crop to a
specific region of interest or unify sources from a few neighboring regions.
These types of operations are the bread and butter of `sf` and `terra`. For
example, if we want to crop the glacier labels down to the region contained by
the satellite image, we can use `sf`’s `st_crop` command.

1. We can also zoom into the raster data using `terra`’s crop function. We crop
both datasets again to get a high-resolution view of this region.
    
    ```{r}
    e <- ext(80.7, 81, 29.9, 30)
    im_crop <- crop(im, e)
    glaciers_crop <- st_crop(glaciers, e)
    tm_shape(im_crop) +
      tm_rgb() +
      tm_shape(glaciers_crop) +
      tm_polygons(col = "Thickness", palette = "Blues", legend.hist = TRUE, alpha = 0.5) +
      tm_layout(legend.outside = TRUE)
    ```
  
1. `tmap` provides many functions for customizing the appearance of our maps.
For example, there are layers for adding specific cartographic elements, like
bar scales (`tmap_scale_bar`) and compasses (`tmap_compass`). Moreover, the themes and
layout of the existing layers can be adapted using the `tmap_layout` function --
this is how we customized the fill colors and legend properties above.
