---
title: Graphical Queries - Brush Events
layout: post
output: 
  md_document:
    preserve_yaml: true
---

_An introduction to brush events in Shiny_

[Code](https://github.com/krisrs1128/stat679_code/blob/main/notes/week3-2.Rmd), [Recording](https://mediaspace.wisc.edu/media/Week%203%20-%202%3A%20Graphical%20Queries%20-%20Brush%20Events/1_y2dyjzbd)

```{r, echo = FALSE}
library(knitr)
opts_knit$set(base_dir = "/", base.url = "/")
opts_chunk$set(
  warning = FALSE, 
  message = FALSE,
  fig.path = "stat679_notes/assets/week3-2/"
  )
```

1. Click events are useful for referring to individual samples. However, they
are not ideal for referring to groups of samples. In this case, a useful type of
plot input is a `brush`. This is a selection that can be defined by clicking and
dragging over a region.

1. In shiny, brush events are treated similarly to click events. For example, to
define a new brush input, we can set the `brush` argument to `plotOutput`.

    ```{r, eval = FALSE}
    ui <- fluidPage(
      plotOutput("plot", brush = "plot_brush")
    )
    ```
  
    Just like the `click` argument, the value `"plot_brush"` is an ID that can be
    used in the server. Also like in click events, we can setup an observer to
    change a reactive value every time a brush is drawn^[Technically, the code
    only executes when the mouse lifts off the brush selection. Some
    visualizations will be able to call the updating code every time the mouse is
    moved with the brush selected. This creates a smoother experience.]. The
    general pattern is similar to what we had before,
      
    ```{r, eval = FALSE}
    server <- function(input, output) {
      selected <- reactiveVal(initial value)
      observeEvent(
        input$plot_brush,
        ... computation using get new_value ...
        selected(new_value)
      )
      
      output$plot <- renderPlot(... use scatter() reactive val...)
    }
    ```
    
1. The example below is similar to the `plot_click` example from the previous
notes. Instead of sorting points by proximity to the click, though, prints the
subset of rows that have been currently brushed.

    ```{r code=readLines("https://raw.githubusercontent.com/krisrs1128/stat479_s22/main/_posts/2022-02-16-week05-02/apps/app3.R"), eval = FALSE}
    ```
    
    <iframe src="https://data-viz.it.wisc.edu/content/a651bc7b-6773-445d-8bc6-22084155e21a/" allowfullscreen="" data-external="1" height=700 width=600></iframe>

1. It is often useful to combine multi-view composition (_i.e._, faceting or
compound figures) with dynamic queries. The basic idea is to (a) show different
aspects of a dataset using different views, and then (b) link the views using
dynamic queries. This strategy is sometimes called dynamic linking.

1. The example below implements dynamic linking with the penguins dataset.
Brushing over either scatterplot highlights the corresponding points in the
adjacent plot (it also updates the data table). This is a way of understanding
structure beyond two dimensions. The implementation is similar to the brushing
above, except that the reactive value `selected()` is called in two `renderPlot`
contexts, leading to changes in both plots every time the brush is moved.

    ```{r code=readLines("https://raw.githubusercontent.com/krisrs1128/stat479_s22/main/_posts/2022-02-16-week05-02/apps/app4.R"), eval = FALSE}
    ```
    
    <iframe src="https://data-viz.it.wisc.edu/content/8e4b07a5-d671-40cb-9d57-22d25533b8fe/" allowfullscreen="" data-external="1" height=800 width=600></iframe>
    
