---
title: Shiny Themes 
layout: post
output: 
  md_document:
    preserve_yaml: true
---

_Custom styling in Shiny._

[Code](https://github.com/krisrs1128/stat679_code/blob/main/notes/week3-4.Rmd), [Recording](https://mediaspace.wisc.edu/media/Week+3+-+4A+Shiny+Themes/1_9ra67tir)

```{r, echo = FALSE}
library(knitr)
opts_knit$set(base_dir = "/", base.url = "/")
opts_chunk$set(
  warning = FALSE, 
  message = FALSE,
  fig.path = "stat679_notes/assets/week3-4/"
  )
```

1. Even if each of its component visualizations were beautifully designed, if we
stick to using the default Shiny styling, it is hard to make our work look
professional. In this set of notes, we’ll discuss how to customize the overall
appearance of a Shiny app using the `bslib` library. We'll work with the Old
Faithful dataset. I've added some filler text to the bottom of the app so that
we can see what effect alternative stylings will have on it.

    ```{r, eval = FALSE}
    library(shiny)
    
    ui <- fluidPage(
      titlePanel("Old Faithful Geyser Data"),
      sidebarLayout(
        sidebarPanel(
          sliderInput("bins", "Number of bins:", min = 1, max = 50, value = 30) 
          ),
        mainPanel(
          plotOutput("distPlot")
          )
      )
    )
    
    server <- function(input, output) {
        output$distPlot <- renderPlot({
            x <- faithful[, 2]
            bins <- seq(min(x), max(x), length.out = input$bins + 1)
            hist(x, breaks = bins, col = 'darkgray', border = 'white',
                 xlab = 'Waiting time to next eruption (in mins)',
                 main = 'Histogram of waiting times')
        })
    }
    
    shinyApp(ui, server)
    ```
      
    <iframe src="https://data-viz.it.wisc.edu/content/b6f80d50-7772-452c-aa09-9320ff3ac811/" allowfullscreen="" data-external="1" height=700 width=600></iframe>

1. `bslib` provides an R interface to bootstrap, which is a web library with
pre-defined website themes. Each theme has been designed in such a way that its
color palettes, font choices, and UI elements (e.g., the shape of buttons) work
well together.

1. The main function in this library is `bs_theme`. The simplest way to
customize a page’s theme is to set the `bootswatch` argument of `bs_theme`,
which globally changes the theme to reflect one of a set of predefined
`bootswatch` themes — you can see choices [here](https://bootswatch.com/).

    ```{r, eval = FALSE}
    library(bslib)
    
    ui <- fluidPage(
      theme = bs_theme(bootswatch = "simplex"),
      titlePanel("Old Faithful Geyser Data"),
      ...
    ```

    <iframe src="https://data-viz.it.wisc.edu/content/0d42915b-1fe1-43f3-81c6-4b47cbb7e23c/" allowfullscreen="" data-external="1" height=700 width=600></iframe>

1. Sometimes we want finer-grained control than simply applying a `bootswatch`
theme. Fortunately, `bs_theme` also provides arguments that allow us to
customize a variety of options.

1. For example, if we want to change the background and foreground colors of our
application, we can use the `bg` and `fg` arguments. Note that I also passed in
a `bg = NA` argument to `renderPlot`, so that the plot was printed
transparently, and not with a white background.

    ```{r, eval = FALSE}
    ui <- fluidPage(
      theme = bs_theme(
        bootswatch = "simplex",
        fg =  "#5DA668",
        bg = "#F2E9E9"
      ),
    ...
    
    output$distPlot <- renderPlot({
      ...
    }, bg = NA)
    ```
    
    <iframe src="https://data-viz.it.wisc.edu/content/8144e782-028e-4c0a-b0a0-3304caced9b9/" allowfullscreen="" data-external="1" height=700 width=600></iframe>
	
1. Similarly, if we want to change the fonts used for headers, main text, and
code, we can set the corresponding themer arguments. It’s possible to define
fonts directly from font objects, but it’s usually easiest to use an existing
link with [Google fonts](https://fonts.google.com/).

    ```{r, eval = FALSE}
    ui <- fluidPage(
      theme = bs_theme(
        bootswatch = "simplex",
        base_font = font_google("Combo"),
        heading_font = font_google("Rubik Moonrocks")
      ),
    ```
    
    <iframe src="https://data-viz.it.wisc.edu/content/e37843db-0cf7-43aa-b204-e26e1fddfa08/" allowfullscreen="" data-external="1" height=700 width=600></iframe>
    
1. A nice trick is that we can customize the appearance of our site
interactively by using the `bs_theme_preview()` function. It will open up our
app with menus for modifying app appearance in real-time. Once we choose a
styling that we like, we can retrieve the corresponding `bs_theme()` arguments
from the R console.

1. Just to illustrate the power of bslib, here is an egregious application of
bslib to create an NES-themed Shiny interface.

    <iframe src="https://data-viz.it.wisc.edu/content/8942c629-fad1-433d-a124-cb56f09b21db/" allowfullscreen="" data-external="1" height=700 width=600></iframe>
