---
title: Manipulating Graph Data
layout: post
output:
  md_document:
    preserve_yaml: true
---

_Reading and modifying graph data_

[Code](https://github.com/krisrs1128/stat679_code/blob/main/notes/week9-1.Rmd), [Recording](https://mediaspace.wisc.edu/media/Week%209%20-%201%3A%20Manipulating%20Graph%20Data/1_co3nvevc)

```{r, echo = FALSE}
library(knitr)
opts_knit$set(base_dir = "/", base.url = "/")
opts_chunk$set(
  cache = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.path = "stat679_notes/assets/week9-1/"
  )
```

```{r}
library(tidygraph)
library(ggraph)
theme_set(theme_bw())
```


1. Before diving into graph data visualization, let’s get some experience
manipulating graphs hands on. One of the best R packages for graph manipulation
is called `tidygraph`. The goal of this package is to extend the semantics of the
tidyverse to graph-structured data. We can't simply use the standard `dplyr`
functions because graphs cannot be stored in simple data.frames -- any graph
must be represented by two data structures, a set of nodes and a set of edges.

1. This can be usefully organized as a pair of `data.frames`, and the
`tidygraph` structure represents graphs in exactly this way.
For example, `G` below is a tidy graph structure showing the friendship
connections between 70 students in a high school over two years. The `Node Data`
component gives features for each student (in this case, just the name), while
`Edge Data` represents friendship links between students.

    ```{r}
    G <- as_tbl_graph(highschool)
    G
    ```

1. The beauty of this data structure is that we can define the analogs of the
usual tidyverse verbs for it. For example, we can derive a new node attribute
using `mutate`.

    ```{r}
    G %>%
      mutate(favorite_color = sample(c("red", "blue"), n(), replace = TRUE))
    ```

1. What if we want to mutate the edges instead? We have to tell tidygraph to
“activate” the edge set,

    ```{r}
    G %>%
      activate(edges) %>%
      mutate(weight = runif(n()))
    ```

    To avoid these activate calls, a convenient shorthand is calling mutate with
    `%N>%` and `%E>%` for modifying node and edge data, respectively,

    ```{r}
    G %E>%
      mutate(weight = runif(n()))
    ```

1. There are many other verbs that have been defined for tidygraph objects. For
example, we can join two graphs together.

    ```{r}
    ## initialize two simple graphs
    G1 <- create_ring(10)  %N>%
      mutate(id = LETTERS[1:n()])
    G2 <- create_bipartite(4, 2) %>%
      mutate(id = LETTERS[1:n()])

    ## join them together
    G1 %>%
      graph_join(G2)
    ```

    ```{r, echo = FALSE}
    library(patchwork)
    p1 <- ggraph(G1) +
      geom_edge_link() +
      geom_node_label(aes(label = id))

    p2 <- ggraph(G2) +
      geom_edge_link() +
      geom_node_label(aes(label = id))

    p3 <- ggraph(G1 %>% graph_join(G2)) +
      geom_edge_link() +
      geom_node_label(aes(label = id))

    p1 | p2 | p3
    ```

1. Similarly, we can filter nodes or edges based on their attributes.

    ```{r}
    G %E>%
      mutate(weight = runif(n())) %>%
      filter(weight < 0.2) %>%
      arrange(-weight)
    ```

1. It’s possible to perform simple graph algorithms using these verbs. For
example, we can cluster nodes based on their connection structure.

    ```{r}
    G %>%
      to_undirected() %>%
      mutate(cluster = group_louvain())
    ```


    ```{r, echo = FALSE}
    G %>%
      to_undirected() %>%
      mutate(cluster = group_louvain()) %>%
      ggraph() +
      geom_edge_link(width = 0.2) +
      geom_node_point(aes(col = as.factor(cluster)), size = 3) +
      scale_color_brewer(palette = "Set2")
    ```

1. We can even map over nodes to compute topological queries. For example, the
block below computes the number of neighbors within two steps of each node,
using the [`local_size` function](https://tidygraph.data-imaginist.com/reference/local_graph.html). More general operations can be computed using map operations, like
[`map_local`](https://tidygraph.data-imaginist.com/reference/map_local.html) or [`map_bfs`](https://tidygraph.data-imaginist.com/reference/map_bfs.html).

    ```{r}
    G %>%
      mutate(two_steps = local_size(order = 2))
    ```
